---
title: "Fake Jira"
author: "Caroline Ring"
date: "2022-09-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Since we can't use real Jira to file tickets forthings to fix, here is a fake Jira

# TCR-001

Entities seem to get dropped when tree is pruned to a certain taxonomy level?

```{r}
seem3path_class <- read.csv("C:\\Users\\cring\\OneDrive - Environmental Protection Agency (EPA)\\treecompareR work\\seem3_path_chems_classified.csv")

org_tree <- prune_tree(chemont_tree, prune_to = "Organic compounds")

#if we prune to class, we see only a few
prune_and_display_subtree(base_tree = org_tree,
                          prune_to = "class",
                          data_1 = seem3path_class,
                          subtree_mapping = list(
                            color= c("gray80",
                                     RColorBrewer::brewer.pal(n=3,
                                                              name = "Dark2")[1]
                                     ),
                            size = c(0.5,1)
                            )
                          )
#looks like what, 9 branches appear?

#but number of unique classes is greater than 9 -- there are 32
length(unique(seem3path_class$class))

#if we prune to subclass we see more
prune_and_display_subtree(base_tree = org_tree,
                          prune_to = "subclass",
                          data_1 = seem3path_class,
                          subtree_mapping = list(
                            color= c("gray80",
                                     RColorBrewer::brewer.pal(n=3,
                                                              name = "Dark2")[1]
                                     ),
                            size = c(0.5,1)
                            )
                          )

#this looks like it's showing closer to 32 branches
#maybe pruning to a taxonomy level isn't working correctly?


```

```{r}
seem3path_tree <- prune_tree(chemont_tree, prune_to = seem3path_class)

prune_and_display_subtree(seem3path_tree, prune_to = "class")
prune_and_display_subtree(seem3path_tree, prune_to = "subclass")
```

```{r}
data_1_nodes <- get_subtree_nodes(data = seem3path_class,
base_tree = org_tree,
tax_level_labels = chemont_tax_levels)

length(data_1_nodes)

#if we take a tree pruned to class as base
data_1_nodes_class <- get_subtree_nodes(data = seem3path_class,
base_tree = prune_tree(org_tree, prune_to = "class"),
tax_level_labels = chemont_tax_levels)
length(data_1_nodes_class)

#if we take a tree pruned to subclass as base
data_1_nodes_subclass <- get_subtree_nodes(data = seem3path_class,
base_tree = prune_tree(org_tree, prune_to = "subclass"),
tax_level_labels = chemont_tax_levels)
length(data_1_nodes_subclass)
```

it's because of this line in `get_subtree_nodes`:

```{r}
 #get labels represented by the classified dataset
  data_labels <- get_terminal_labels(data = data,
                                     tax_level_labels = tax_level_labels)
#get node numbers, levels, & names of base tree
  tree_df <- get_tree_df(base_tree)

  #get node numbers represented in dataset
  data_nodes <- tree_df[tree_df$Name %in% data_labels, "node"]
```

It looks for *terminal* labels of the data in the base tree, which means that when the base tree is pruned to a certain level, it will only keep entities whose *terminal* label was at that level.

Instead we need to handle it a little bit differently.

I think we need to take all the data labels, not just the terminal ones.

OK -- fixed as of 2022-09-08 12:10 PM by Caroline Ring commit SHA abac8b0c

# TCR 101

In display_subtree, if data_2 is given as a phylo tree, it will not be highlighted. However, it works fine if the tree is given as data_1.

Fixed with commit SHA f65c72fc6
Caroline Ring 2022-09-08 03:21 PM

# TCR 102

Bug in classify_structures() around line 384
When trying to run on first 10 SMILES in biosolids_class, get the folowing error:
Error in (function (..., row.names = NULL, check.rows = FALSE, check.names = TRUE,  : 
  arguments imply differing number of rows: 0, 1
  
Presumably this arises when there are no missing items



